Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE)

### 2014-02-06 WSSSPE1_report/WSSSPE1_summary_report.tex
Update WSSSPE1_summary_report.tex

Additional volunteers:
\begin{itemize}
\item Chris Mattmann $\langle$\url{chris.a.mattmann@jpl.nasa.gov}$\rangle$
\item * Ketan Maheshwari $\langle$\url{ketancmaheshwari@gmail.com}$\rangle$
\item Marlon Pierce $\langle$\url{marpierc@iu.edu}$\rangle$
\item Colin Venters $\langle$\url{C.Venters@hud.ac.uk}$\rangle$
\item Suresh Marru $\langle$\url{smarru@iu.edu}$\rangle$
\item Lynn Zentner $\langle$\url{lzentner@purdue.edu}$\rangle$
\item Anne Elster $\langle$\url{anne.elster@gmail.com}$\rangle$
\item Shel Swenson $\langle$\url{mdswenso@usc.edu}$\rangle$
\item Andy Ray Terrel $\langle$\url{andy.terrel@gmail.com}$\rangle$
\item Abani Patra $\langle$\url{abani.patra@gmail.com}$\rangle$
\item Nancy Wilkins-Diehr $\langle$\url{wilkinsn@sdsc.edu}$\rangle$
\item James Spencer $\langle$\url{j.spencer@imperial.ac.uk}$\rangle$
\item * Frank L\"{o}ffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
\item * James Hetherington $\langle$\url{j.hetherington@ucl.ac.uk}$\rangle$
\item * Sou-Cheng (Terrya) Choi $\langle$\url{sou.cheng.terrya.choi@gmail.com}$\rangle$
\item James Howison $\langle$\url{james@howison.name}$\rangle$
\item * Bruce Berriman $\langle$\url{gbb@ipac.caltech.edu}$\rangle$
\item Hilmar Lapp $\langle$\url{hlapp@nescent.org}$\rangle$
\item * Marcus D. Hanwell $\langle$\url{marcus.hanwell@kitware.com}$\rangle$
\item Lucas Nussbaum $\langle$\url{lucas.nussbaum@loria.fr}$\rangle$
\item Matthew Turk $\langle$\url{matthewturk@gmail.com}$\rangle$
\end{itemize}

@@

The First Workshop on Sustainable Software for Science: Practice and
Experiences (WSSSPE1,
\url{http://wssspe.researchcomputing.org.uk/WSSSPE1}) was held on
Sunday, 17 November 2013, in conjunction with the 2013 International
Conference for High Performance Computing, Networking, Storage and
Analysis (SC13, \url{http://sc13.supercomputing.org}).

Because progress in scientific research is dependent on the quality
and accessibility of software at all levels, it is now critical to
address many challenges related to the development, deployment, and
maintenance of reusable software. In addition, it is essential that
scientists, researchers, and students are able to learn and adopt
software-related skills and methodologies. Established researchers are
already acquiring some of these skills, and in particular a
specialized class of software developers is emerging in academic
environments who are an integral and embedded part of successful
research teams. The WSSSPE workshop was intended to provide a forum
for discussion of the challenges, including both positions and
experiences.  The short papers and discussion have been archived to
provide a basis for continued discussion, and the workshop has fed
into the collaborative writing of this journal publication.
Additional papers based on extended versions of workshop submissions
are expected.  The level of interest in the workshop has led the
organizers, working with some of the submitters and attendees, to plan
two additional workshops (at the 2014 SciPy and SC conferences) and to
turn the workshop website into a community website that can be used as
a focus for further discussion and progress.  A minisymposium on
``Reliable Computational Science'' that aims to further explore some
of the key issues raised in WSSSPE is co-organized by a participant in
the 2014 SIAM Annual Meeting.

Before the WSSSPE1 workshop took place, the organizers self-published
a paper~\cite{WSSSPE1-pre-report} to document the process of
organizing and advertising the workshop, collecting and reviewing the
papers, and putting together the agenda.  Section~\ref{sec:process} is
based on that paper.  The remainder of this paper is based on the
workshop itself, as documented in a set of collaborative notes taken
during the workshop~\cite{WSSSPE1-google-notes}, including discussion
about the keynote presentations (\S\ref{sec:keynotes}), the three
panels (\S\ref{sec:devel}-\ref{sec:community}), other discussion
(\S\ref{sec:other}), cross-cutting issues (\S\ref{sec:cross-cutting}),
use cases (\S\ref{sec:use-cases}), and conclusions
(\S\ref{sec:conclusions}).

\todo{Mention somewhere statistics of participants: There were 140 seats, I (Katz) think 149 people at the first keynote, and about 90 at the end of the workshop.  They (SC13) counted people in the room 4 times during the day.  But I don’t know how many unique people attended, which I agree they should know.}

\section{Workshop Process and Agenda} \label{sec:process}

WSSSPE1 was organized in collaboration of the relatively small group
of five organizers and a larger peer-review committee with 36
members. This committee had early influence on the organization, e.g.,
on the specific call for papers.

@@

\begin{itemize}[leftmargin=0.2in]
\item the development process that leads to new (versions of) software
\begin{itemize}[leftmargin=0.2in]
\item how fundamental research in computer science or
  science/engineering domains is turned into reusable software
\item software created as a by-product of research
\item impact of computer science research on the development of
  scientific software
\end{itemize}
\item the support and maintenance of existing software, including
  software engineering
\begin{itemize}[leftmargin=0.2in]
\item governance, business, and sustainability models
\item the role of community software repositories, their operation and
  sustainability
\end{itemize}
\item the role of open source communities or industry
\item use of the software
\begin{itemize}[leftmargin=0.2in]
\item growing communities
\item reproducibility, transparency needs that may be unique to science
\end{itemize}
\item policy issues, such as
\begin{itemize}[leftmargin=0.2in]
\item measuring usage and impact
\item software credit, attribution, incentive, and reward
\item career paths for developers and institutional roles
\item issues related to multiple organizations and multiple countries,
  such as intellectual property, licensing, etc.
\item mechanisms and venues for publishing software, and the role of
  publishers
\end{itemize}
\item education and training
\end{itemize}

@@

58 submissions were received, and almost all submitters used either
arXiv~\cite{arXiv-web} or figshare~\cite{figshare-web} to self-publish
their papers.

A peer review process followed the submissions, where
the 58 papers received 181 reviews, an average of 3.12 reviews per
paper. Reviews were completed using a Google form, which allowed
reviewers to choose papers they wanted to review, then to provide
general comments and scores on relevance to the organizers and
to the authors. The review reports enabled the organizers to decide
which papers to associate with the workshop, and allowed the authors
to improve their papers.

The organizers decided to list 54 of the papers as significant
contributions to the workshop, a very high acceptance rate, but one
that is reasonable, given the goal of broad participation and the fact
that the reports were already self-published. The papers were
grouped into three main categories, namely \emph{Developing and
Supporting Software}, \emph{Policy}, and \emph{Communities}. Each
subject area was associated with a panel and discussion at the
workshop.

The workshop itself consisted of two keynote presentations and the
three panels/discussion sessions. The panels were organized based on a
classification of the workshop submissions into three categories,
following the themes of the call for papers as modified by the areas
of the submissions. Three to four representatives from each submission
category were appointed as panelists, and assigned to read a subset of
the paper in that category and then discuss them in the panel.

\section{Keynotes} \label{sec:keynotes}

\note{lead: Sou-Cheng (Terrya) Choi}

The WSSSPE1 workshop began with two keynote presentations.

@@

The first
keynote\footnote{\url{http://wssspe.researchcomputing.org.uk/keynotes/\#bourne}}
was delivered by Philip E. Bourne of UCSD. The slides are available
online.\footnote{\url{www.slideshare.net/pebourne/a-recipe-for-sustainable-software}}
Philip is a basic biomedical scientist but has formed four software
companies. He also helped found PLOS and
PDB\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.  He is
working on automating three-dimensional visualizations of cell
contents and molecular structures, a problem that has not been solved
and would serve as a key function of software in biomedical sciences.

Philip emphasized that Sustainability for software ``does not just
mean more money from Government.''  He noted that in places,
development in science has improved thanks to open source and Github,
but for the most part remains arcane. He argued that there is much we
can learn from the app-store model. He mentioned
BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}} and
OpenScienceDataCloud\footnote{\url{https://www.opensciencedatacloud.org/}}
as distribution channels.  On a related note, Philip observed a common
evolutionary pathway for computational biology projects, from data
archive to analytics platform to educational use, and suggested that
use of scientific software for outreach might be the final step.

Philip shared with the audience a few real challenges he
encountered. First, he has looked into reproducibility in
computational biology, and has concluded that ``I have proved I cannot
reproduce research from my own lab''~\cite{Veretnik}.

Another problem he experienced was staff retention with respect to
private organizations which reward those combining research and
software expertise (the ``Google Bus''). However, he is a strong fan
of private sector sustainability sources and public-private
partnerships. He noted that making a successful business from
scientific software alone is very rare: founders overvalue while
customers undervalue. He noted that to last, an open source project
needs a minimal funding requirement even with a vibrant community --
good will only goes so far if you are being paid to do something else.
He talked about grant schemes of relevance in the U.S., particularly
with regard to technology
transfer.\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}}

He also had problems with selling research software: the technology
transfer office in his university wanted huge IP reachthrough, whereby
they would get a share of profits from drugs developed by pharma
companies who use the software. He was aware this was unrealistic but
the technology transfer office insisted for a while. He wants to push
a one-click approach to purchasing university-written software.

@@

On the idea of the \textbf{digital enterprise}, where
IT underpins the whole of organizational activities, he contended that
universities are way behind the curve on this. In particular, he
highlighted the separation of research, teaching and administration into
silos without a common IT framework as a blocker to many useful
organizational innovations: ``University 2.0 is yet to happen.'' He
spoke of a circumstance where someone had used an algorithm developed
for computational biology in marketing.

@@

\section{Developing and Supporting Software} \label{sec:devel}

\note{lead: Marcus Hanwell, contributors: Suresh Marru}

@@

\section{Policy} \label{sec:policy}

\note{lead: Colin Venters}

@@

\section{Communities} \label{sec:community}

\note{lead: Matthew Turk, contributors: Nancy Wilkins-Diehr, Chris Mattmann, Suresh Marru, Frank L\"{o}ffler, Andy Terrel}

@@

@@

\note{all to put notes here of stuff that doesn't fit elsewhere, if any}

\todo{perhaps remove this section}

\section{Cross-cutting Issue: Defining Sustainability}  \label{sec:cross-cutting}

\note{lead: Daniel S. Katz}

The question of what was meant by ``sustainability'' in the context of software
came up in many different parts of the workshop, specifically in one of the
keynotes, the Developing and Maintaining Software panel, and the Policy panel. \todo{check closing discussion notes for this topic}

In his keynote, Phil Bourne suggested that perhaps sustainability can be defined
as the effort that happens to make the essential things continue. This leads to
having to decide what it is that we want to sustain, if what we want to sustain is
valuable, and finally, who would care if it went away, and how does one measure
how much they care.

\note{rest of section will be turned from bullets into text...}

In the Developing and Maintaining Software panel,

\begin{itemize}
\item what does sustainable mean?
\begin{itemize}
\item from http://oss-watch.ac.uk/resources/ssmm: When choosing
  software for procurement or development reuse - regardless of the
  license and development model you will use - you need to consider
  the future. While a software product may satisfy today's needs, will
  it satisfy tomorrow's needs? Will the supplier still be around in
  five years' time? Will the supplier still care for all its customers
  in five years' time? Will the supplier be responsive to bug reports
  and feature requests? In other words, is the software sustainable?
\item I think this is a key question we as a community need to agree
  on (whats our problem statement?) and hope we come out today with
  some progress. - @VonWelch
\item What is the goal of sustainability? Reproducible science,
  persistence, quality, or else?
\item How do you measure success? How do you know when you have
  actually achieved sustainable software?
\item Sustainability should be addressed throughout the life cycle
  (Daisie Huang)
\end{itemize}

\item What is the relationship between sustainability and provenance?

\item Is usability separate from sustainability or a fundamental part?
\end{itemize}

In the Policy panel,

\begin{itemize}
\item Chris Lenhardt's presentation of Venters
  paper~\cite{Venters_WSSSPE} included:

\begin{itemize}
\item What does software sustainability mean
It ``is a rather ambiguous concept...''
\item Lack of an accepted definition gets in the way of integrating
  the concept into software engineering
\item Sustainability as a non-functional requirement
\item Quality of software architectures determines sustainability
\item Propose sustainability is a measure of central quality
  attributes: Extensibility, Interoperability, Maintainability,
  Portability, Reusability, Scalability
\item Develop an architecture evaluation framework based on scenarios
  which help to illuminate how to measure quality/sustainability at
  the architectural level
\end{itemize}

\item Chris Lenhardt's presentation of Katz paper~\cite{Katz_WSSSPE}
  included:
\begin{itemize}
\item Clarifying the notion of sustainability
\begin{itemize}
\item Provide same functionality going forward even if environment changes
\item Is functionality and usability clear
\item Will functionality be correct going into the future even if
  environment changes
\item Is the functionality desired by current and future users
\item Can it incorporate new things as they develop
\end{itemize}
\end{itemize}

\item see slide 9 of
  \url{http://dx.doi.org/10.6084/m9.figshare.853817} for summary table

\item also see discussion at \url{https://docs.google.com/document/d/1eVfioGNlihXG_1Y8BgdCI6tXZKrybZgz5XuQHjT1oKU/edit?pli=1#bookmark=id.2pxw5yey8jl0}

\section{Illustrative Use Cases} \label{sec:use-cases}

\note{lead: Ketan Maheshwari, contributors: Andy Terrel}

@@

\note{Dan: do we need this section?  I think there's some overlap
  between this and the discussion in the Developing and Support
  Software section.  Or perhaps this should go in the cross-cutting
  issues section?  Let's write it down here - then we can decide if it goes elsewhere}

\section{Conclusions} \label{sec:conclusions}

@@

\todo{anyone who needs to put something in here should}

Some of the work by Katz was
supported by the National Science Foundation while working at the
Foundation; any opinion, finding, and conclusions or recommendations
expressed in this material are those of the author(s) and do not
necessarily reflect the views of the National Science Foundation.

@@

\note{lead: Shel Swenson}

\note{should alphabetize all, but if we get more from the SC13 folks, it's not worth spending
time doing this now.  I wonder if there's a way to just have a simple list that we can keep ordered, and let LaTeX do the formatting, rather than manually creating a table as this is now}

@@

\begin{tabular}{ L{4cm}  L{4cm}  L{4cm}  L{4cm} }
   Daniel S. Katz
&  Neil Chue Hong
&  Gabrielle Allen
&  David Proctor
\\ Manish Parashar
&  Jay Alameda
&  David Andrs
&  Lorena A Barba
\\ David Bernholdt
&  Phil Bourne
&  Karl Broman
&  Maxine Brown
\\ David Bruhwiler
&  Bruno Bzeznik
&  Jeffrey Carver
&  John W. Cobb
\\ Karen Cranston
&  Rion Dooley
&  Anshu Dubey
&  Ian Foster
\\Juliana Freire
& Derek Gaston
& Derek Groen
& Edmund Hart
\\James Hetherington
& Simon Hettrick
& Kenneth Hoste
& James Howison
\\Daisie Huang
& Kerk Kee
& Arnaud Legrand
& Paul Madden
\\Louis Maddox
& Ketan Maheshwari
& Suresh Marru
& Cezary Mazurek
\\James McClure
& Chris Mentzel
& J. Yates Monteith
& Pierre Neyron
\\Lucas Nussbaum
& Cody Permann
& John Peterson
& Quan Pham
\\Marlon Pierce
& Heather Piwowar
& Jason Riedy
& Andreas Schreiber
\\William Scullin
& Andrew Slaughter
& Arfon Smith
& James Spencer
\\Shel Swenson
& Fr\'{e}d\'{e}ric Suter
& Andy Terrel
& George  Thiruvathukal
\\John Towns
& Erik Trainer
& Matthew Turk
& Eric Van Wyk
\\Colin C. Venters
& Von Welch
& Nancy Wilkins-Diehr
& Theresa Windus
\\Rich Wolski
& Lynn Zentner
& Frank L\"{o}ffler
& Patrick O'Leary
\end{tabular}

>>>>

Additional volunteers:
\begin{itemize}
\item Chris Mattmann $\langle$\url{chris.a.mattmann@jpl.nasa.gov}$\rangle$
\item * Ketan Maheshwari $\langle$\url{ketancmaheshwari@gmail.com}$\rangle$
\item Marlon Pierce $\langle$\url{marpierc@iu.edu}$\rangle$
\item Colin Venters $\langle$\url{C.Venters@hud.ac.uk}$\rangle$
\item Suresh Marru $\langle$\url{smarru@iu.edu}$\rangle$
\item Lynn Zentner $\langle$\url{lzentner@purdue.edu}$\rangle$
\item Anne Elster $\langle$\url{anne.elster@gmail.com}$\rangle$
\item Shel Swenson $\langle$\url{mdswenso@usc.edu}$\rangle$
\item Andy Ray Terrel $\langle$\url{andy.terrel@gmail.com}$\rangle$
\item Abani Patra $\langle$\url{abani.patra@gmail.com}$\rangle$
\item Nancy Wilkins-Diehr $\langle$\url{wilkinsn@sdsc.edu}$\rangle$
\item James Spencer $\langle$\url{j.spencer@imperial.ac.uk}$\rangle$
\item * Frank Löffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
\item * James Hetherington $\langle$\url{j.hetherington@ucl.ac.uk}$\rangle$
\item * Sou-Cheng (Terrya) Choi $\langle$\url{sou.cheng.terrya.choi@gmail.com}$\rangle$
\item James Howison $\langle$\url{james@howison.name}$\rangle$
\item * Bruce Berriman $\langle$\url{gbb@ipac.caltech.edu}$\rangle$
\item Hilmar Lapp $\langle$\url{hlapp@nescent.org}$\rangle$
\item * Marcus D. Hanwell $\langle$\url{marcus.hanwell@kitware.com}$\rangle$
\item Lucas Nussbaum $\langle$\url{lucas.nussbaum@loria.fr}$\rangle$
\item Matthew Turk $\langle$\url{matthewturk@gmail.com}$\rangle$
\end{itemize}

@@

The First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)\footnote{\url{http://wssspe.researchcomputing.org.uk/WSSSPE1}} was held on Sunday, 17 November 2013, in conjunction with the 2013 International Conference for High Performance Computing, Networking, Storage and Analysis (SC13)\footnote{\url{http://sc13.supercomputing.org}}.

Because progress in scientific research is dependent on the quality and accessibility
of software at all levels, it is now critical to address many challenges related to the
development, deployment, and maintenance of reusable software. In addition, it is
essential that scientists, researchers, and students are able to learn and adopt
software-related skills and methodologies. Established researchers are already
acquiring some of these skills, and in particular a specialized class of software
developers is emerging in academic environments who are an integral and embedded
part of successful research teams. The WSSSPE workshop was intended to provide a
forum for discussion of the challenges, including both positions and experiences.
The short papers and discussion have been archived to provide a basis for continued
discussion, and the workshop has fed into the collaborative writing of this journal publication.
Additional papers based on extended versions of workshop submissions are expected.
The level of interest in the workshop has led the organizers, working with some of the submitters
and some of the attendees, to plan two additional workshops (\todo{details}) and to turn
the workshop website into a community website that can be used as a focus for further
discussion and progress.

\todo{Mention a two-session minisymposium on ``Reliable Computational Science'' in the SIAM Annual Meeting in July. It aims to further explore some of the key issues raised in WSSSPE. Speakers include Tim Davis, Barry Smith, Lorena Barba, Tanu Malik, Bill Rider.}

\todo{mention arXiv:1311.3523 and explain what's different here}

\section{Workshop Process}

WSSSPE1 was organized in collaboration of the relatively small group of \todo{how many} organizers and
a larger peer-review committee consisting of \todo{how many members}. This committee had early influence on the organization,
e.g., on the specific call for papers.

@@

\begin{itemize}
\item the development process that leads to new (versions of) software
\begin{itemize}
\item how fundamental research in computer science or science/engineering domains is turned  into reusable software
\item software created as a by-product of research
\item impact of computer science research on the development of scientific software
\end{itemize}
\item the support and maintenance of existing software, including software engineering
\begin{itemize}
\item governance, business, and sustainability models
\item the role of community software repositories, their operation and sustainability
\end{itemize}
\item the role of open source communities or industry
\item use of the software
\begin{itemize}
\item growing communities
\item reproducibility, transparency needs that may be unique to science
\end{itemize}
\item policy issues, such as
\begin{itemize}
\item measuring usage and impact
\item software credit, attribution, incentive, and reward
\item career paths for developers and institutional roles
\item issues related to multiple organizations and multiple countries, such as intellectual property, licensing, etc.
\item mechanisms and venues for publishing software, and the role of publishers
\end{itemize}
\item education and training
\end{itemize}

@@

58 submissions were received, and almost all submitters used either 
arXiv\footnote{\url{http://arxiv.org}} or 
figshare\footnote{\url{http://figshare.com}} to self-publish their papers.

A peer review process followed the submissions, where the 58 papers received 181 reviews, an 
average of 3.12 reviews per paper. Reviews were completed using a Google form, which allowed 
reviewers to provide scores on relevance and comments to the organizers and to the authors. The review reports enabled the organizers to decide which papers to associate with the workshop, and the authors to improve their papers.

The organizers decided to list 54 of the papers as significantly contributing to the workshop, a very 
high acceptance rate, but one that is reasonable, given the goal of broad participation and the fact 
that the reports were already self-published. The papers were also grouped into three main categories, namely \emph{Developing and Supporting Software}, \emph{Policy}, and \emph{Communities}. Each subject area was associated with a panel and discussion at the workshop.

\section{Agenda}

The workshop consisted of two keynote presentations and three panels/discussion sessions. The panels 
were organized based on a classification of the workshop submissions into three categories, following 
the themes of the call for papers as modified by the areas of the submissions. Four \todo{Is it five?} representatives from 
each submission categories were appointed as panelists, to discuss the papers in that category.

\section{Keynotes}

@@

The first keynote\footnote{\url{http://wssspe.researchcomputing.org.uk/keynotes/\#bourne}}
was delivered by Philip E. Bourne of UCSD. The slides
are online.\footnote{\url{www.slideshare.net/pebourne/a-recipe-for-sustainable-software}}

Philip is a basic biomedical scientist but has formed four software
companies. He also helped found PLOS and PDB\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.

He emphasized that Sustainability for software ``does not just mean more
money from Government.''

He noted that creating effective illustrations is a key function of
software in the biomedical sciences: ``Automating Conceptualization''
(Visualizing molecular structures is a problem that has now been solved,
he is working on 3-d visualizations of the contents of cells.)

He has looked into reproducibility in computational biology, and has
concluded that ``I have proved I cannot reproduce research from my own
lab''~\cite{Veretnik}.

He noted that in places, development in science has improved thanks to
open source and Github, but for the most part remains arcane. He argues
that there is a lot we can learn from the app-store model. He
mentioned BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}} and
OpenScienceDataCloud \footnote{\url{https://www.opensciencedatacloud.org/}} as
distribution channels.

He noted a common evolutionary pathway for computational biology
projects, from data archive to analytics platform to educational use. He
suggests that use for outreach might be the final step.

He experienced problems with staff retention with respect to private organizations which reward 
those combining research and software expertise (the ``Google Bus''). However,
he is a strong fan of private sector sustainability sources and PPP. He
noted that making a successful business from scientific software alone
is very rare: founders overvalue while customers undervalue. He noted
that to last, an open source project needs a minimal funding requirement
even with a vibrant community -- good will only goes so far if you are
being paid to do something else.

He talked about grant schemes of relevance in the U.S., particularly with regard to technology transfer.\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}} \todo{change at least first to cite}

He had problems with selling research software: the technology transfer
office in his university wanted huge IP reachthrough, whereby they would get a
share of profits from drugs developed by pharma companies who use the
software. He was aware this was unrealistic but the technology transfer
office insisted for a
while. He wants to push a one-click approach to purchasing
university-written software.

@@

He then expanded on the idea of the \textbf{digital enterprise}, where
IT underpins the whole of organizational activities. He contends that
universities are way behind the curve on this. In particular, he
highlights the separation of research, teaching and administration into
silos without a common IT framework as a blocker to many useful
organizational innovations: ``University 2.0 is yet to happen.'' He
spoke of a circumstance where someone had used an algorithm developed
for computational biology in marketing.

@@

\section{Developing and Supporting Software}

@@

\section{Policy}

@@

\section{Communities}

@@

There were a number of papers categorized under the Communities banner and so two presenters each summarized half of the submissions in this area.

\subsection{Communities Part 1}

This collection of papers was summarized by Karen Cranston.

\subsection{Communities Part 2}

Nancy Wilkins-Diehr summarized 6 papers in "Communities Part 2". The Christopherson paper outlines the degree to which research
relies on high quality software. There are often barriers and a lack of suitable incentives for researchers to embrace software engineering principals.
The Water Science Software Institute is working to lower some of these barriers through an Open Community Engagement Process.
This is a 4-step iterative development process that incorporates Agile development principals.

\item Step 1 - Design - thorough discussion of research questions
\item Step 2 - Develop working code
\item Step 3 - Refine based on new requirements
\item Step 4 - Publish open source

The Christopherson paper reports on the application of Steps 1-3 to a computational modeling framework developed in the 1990s.
Step 1 was realized as a 2-day, in person specifications meeting and code walk-through. Step 2 was a 5-day hack-athon to develop working code.
Step 3 was a 3-day hack-a-thon to refine the code based on new requirements. The team worked on small, low-risk unit of 
code. It was challenging, revealed unanticipated obstacles, programmers had to work together, and experimentation was encouraged.

The paper summarized recommendations to others wishing to engage in this or a similar process. Start small and gradually 
build toward more complex objectives. This is consistent with Agile development. Refactor before adding new functionality.
Approach development as a learning experience. Welcome experimentation, and treat mistakes as a natural part of the learning process.
Repeat Step 1 activities before all hackathons to develop consensus before coding. This allows coding to be the focus of subsequent hackathons. 
In higher risk situations, provide additional time for Step 1 activities. The Christopherson team recommends a minimum of two months.
Ensure any newcomers receive some form of orientation prior to the hackathon, such as a code walkthrough or system documentation.
Co‐locate rather than collaborating remotely whenever feasible.

The Pierce paper described how science gateways can provide a user-friendly entry to complex cyberinfrastructure. The paper opens with a description of the explosion of use of just a few gateways. 
Over 3.5 years, more than 7,000 biologists have run phylogenetic codes on supercomputers using the CIPRES Science Gateway. In 1 year, over 120 scientists from 50 institutions used the UltraScan Science Gateway, 
increasing the sophistication of analytical ultracentrifugation experiments worldwide. The new Neuroscience Gateway (NSG) registered 100+ 
users who used 250,000 CPU hours in only a few months.

Gateways, however, need to keep operational costs low and can often make use of common components. Science Gateway Platform as a Service (Sci-GaP)
delivers middleware as a hosted, scalable third ­party service while domain developers focus on user interfaces and domain-specific data in the gateway.
The middleware handles things like authentication, application installation and reliable execution and help desk support.

One key to SciGaP is its openness. While based on Apache Airavata project and the CIPRES Workbench Framework, community contributions are encouraged because of its open source, open governance and open operations policies.
The goal is robust, sustainable infrastructure with a cycle of development that improves reliability and prioritizes
stakeholder requirements. The project is leveraging Internet2’s Net+ mechanisms for converting SciGaP and its gateways into commodity services.

The Zentner paper describes experiences and challenges with the large nanoHUB.org community. They define community as 
a “body of persons of common and especially professional interests scattered through a larger society.” This makes support
challenging because of the diversity of viewpoints and needs. The group constantly examines its policies to determine
whether they are indirectly alienating part of the community or encouraging particular types of use.

nanoHUB's 10-year history with over 260,000 users annually provides a lot of data to analyze. 4000 resources contributed by
1000 authors. nanoHUB serves both the research and education community and the contribution model allows researchers to get
their codes out into the community and in use in education very rapidly. The primary software challenges are twofold - support
for the HUBzero framework and challenges related to the software contributed by the community.

The group has learned that community contributions are maximized with a tolerant licensing approach. HUBzero uses an 
LGPLv3 license so contributors can create unique components and license as they choose. If they make changes to source 
code, the original license must be maintained for redistribution. As far as contributed resources, these must be open access,
but not necessarily open source. This allows contributors to meet the requirements of their institutions and funding agencies.
Quality is maintained via user ratings. Documentation is encouraged and nanoHUB supplies regression test capabilities, but 
the user community provides ratings, poses questions and contributes to wishlists and citation counts - all of which incentivize
code authors.

The Terrel paper describes support for the Python scientific community through two major efforts - the SciPy conference and the
NumFOCUS foundation. Reliance on software in science has driven a huge demand for development, but this development is typically done as a side effort
and often in a rush to publish without documentation and testing.
While created by academics, software support often falls to industrial institutions. SciPy brings together industry, government, 
and academics to share their code and experience in an open environment.

NumFOCUS promotes open, usable scientific software while sustaining the community. Specific activities include educational programs, 
collaborative research tools and documentation and 
promotion of high­-level languages, reproducible scientific research, and open­code developmet.
Governance of the non-profit is a loose grantor-­grantee relationship with projects allowing for monies to be placed in the groups accounts.
This has raised money to hire developers for open code development, maintain testing machines, organize the PyData conference series, and sponsor community members to attend conferences.

Software sustainability relies on contributions from all sectors of user community. Together SciPy and NumFOCUS support these sectors. By maximizing contributions growing the user base they help develop and mature Python.

The Loffler paper

@@

\section{Cross-cutting Issues}

\subsection{Defining sustainability}

\subsection{Licensing}

\note{not sure if this was really cross-cutting...}

\section{Illustrative use-cases}

@@

\section{Conclusions}

@@

@@

@@

Daniel S. Katz (@danielskatz)

Neil Chue Hong (@npch)

Gabrielle Allen

David Proctor (@djproctor)

Manish Parashar

Jay Alameda (alameda@illinois.edu)

David Andrs

Lorena A Barba (@LorenaABarba)

David Bernholdt (bernholdtde@ornl.gov)

Phil Bourne

Karl Broman (@kwbroman)

Maxine Brown, UIC (maxine@uic.edu)

David Bruhwiler   (bruhwiler@radiasoft.net    @radiasoft)

Bruno Bzeznik

Jeffrey Carver, University of Alabama (carver@cs.ua.edu)

John W. Cobb (cobbjw@ornl.gov, @johnwcobb)

Karen Cranston (@kcranstn, National Evolutionary Synthesis Center)

Rion Dooley

Anshu Dubey (adubey@lbl.gov)

Ian Foster, @ianfoster

Juliana Freire

Derek Gaston

Derek Groen (@whydoitweet on Twitter)

Edmund Hart (@distribecology)

James Hetherington @jamespjh

Simon Hettrick (s.hettrick@software.ac.uk)

Kenneth Hoste (@kehoste, Ghent University, Belgium)

James Howison, University of Texas at Austin (@jameshowison)

Daisie Huang

Kerk Kee (kerk.kee@gmail.com)

Arnaud Legrand

Paul Madden, NOAA / University of Colorado, maddenp@colorado.edu

Louis Maddox (@lmmddx)

Ketan Maheshwari (@km\_9 on Twitter)

Suresh Marru (smarru@iu.edu, @sureshmarru)

Cezary Mazurek, mazurek@man.poznan.pl, @cezary\_mazurek

James McClure

Chris Mentzel @chrismentzel - firstname dot lastname at moore dot org

J. Yates Monteith, Clemson University

Pierre Neyron

Lucas Nussbaum (lucas.nussbaum@loria.fr)

Cody Permann

John Peterson

Quan Pham, University of Chicago

Marlon Pierce, marpierc@iu.edu, @marpierc

Heather Piwowar

Jason Riedy, Georgia Tech, http://fmrl.me/jasonriedy and many other places

Andreas Schreiber (@onyame)

William Scullin

Andrew Slaughter

Arfon Smith

James Spencer (@exp\_n)

Shel Swenson, University of Southern California (shel.swenson@usc.edu)

Fr\'{e}d\'{e}ric Suter

Andy Terrel (@aterrel, andy.terrel@gmail.com, NumFOCUS and Continuum Analytics)

George K. Thiruvathukal, Loyola University Chicago and Computing in Science and Engineering (IEEE CS/AIP) (gkt@cs.luc.edu)

John Towns

Erik Trainer

Matthew Turk (@powersoffour on Twitter, matthewturk@gmail.com)

Eric Van Wyk, University of Minnesota, evw@cs.umn.edu, @ericvanwyk

Colin C. Venters, c.venters@hud.ac.uk

Von Welch, @VonWelch

Nancy Wilkins-Diehr, wilkinsn@sdsc.edu

Theresa Windus (twindus@iastate.edu)

Rich Wolski

Lynn Zentner

Frank Löffler
